Selection on early survival does not explain germination rate clines in \emph{Mimulus cardinalis} (Phrymaceae)

### 2022-06-23 ms/ms.tex
r1 prerelease

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{setspace}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Selection on early survival does not explain germination rate clines in Mimulus cardinalis (Phrymaceae)},
  pdfauthor={Christopher D. Muir\^{}\{1,2,*\}, Courtney L. Van Den Elzen\^{}\{1,3\}, Amy L. Angert\^{}\{1,4\}},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{geometry}
\geometry{verbose,letterpaper,margin=2.45cm}

@@

\usepackage{lineno}
% \linenumbers % toggle on for linenumbers

@@

\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

@@

\setstretch{1}
\(^{1}\) Department of Botany, University of British Columbia, Vancouver, BC V6T 1Z4, Canada

@@

@@

\textbf{Conclusions} Southern populations have higher early survival but this is not caused by faster germination. This pattern is consistent with nonadaptive forces driving the phenotypic cline in germination rate, but future work will need to assess whether there is selection at other life stages. The statistical framework should help expand quantitative genetic analyses for other waiting-time traits.

@@

Populations within species differ in ecologically important traits that often result from adaptation to different local environments (\protect\hyperlink{ref-turesson_genotypic_1922}{Turesson, 1922}; \protect\hyperlink{ref-clausen_experimental_1948}{Clausen et al., 1948}). However, it is still rare that we understand the traits and selective agents underlying local adaptation (\protect\hyperlink{ref-wadgymar_identifying_2017}{Wadgymar, Lowry, et al., 2017}). The classic signature of local adaptation is crossing reaction norms for fitness measured in a reciprocal transplant experiment (\protect\hyperlink{ref-kawecki_conceptual_2004}{Kawecki and Ebert, 2004}; \protect\hyperlink{ref-johnson_reciprocal_2021}{Johnson et al., 2021}), such that populations have higher relative fitness in their local environment but lower relative fitness in a foreign environment. For example, annual inland ecotypes of the yellow Monkeyflower \emph{Mimulus guttatus} have higher relative fitness in seasonally dry sites compared to perennial coastal ecotypes that are adapted to year-round water availability. The relative fitness of coastal and inland ecotypes is reversed on the coast because of life history and other genetic differences between ecotypes (\protect\hyperlink{ref-lowry_widespread_2010}{Lowry and Willis, 2010}). Local adaptation is important in evolutionary biology because its prevalence implies that selection varies spatially, populations have heritable variation in fitness, and selection is stronger than migration or drift. Exceptional cases of local adaptation over short temporal or spatial scales (e.g. \protect\hyperlink{ref-wright_indirect_2013}{Wright et al.} (\protect\hyperlink{ref-wright_indirect_2013}{2013}); \protect\hyperlink{ref-grant_40_2014}{Grant and Grant} (\protect\hyperlink{ref-grant_40_2014}{2014}); \protect\hyperlink{ref-richardson_microgeographic_2014}{Richardson et al.} (\protect\hyperlink{ref-richardson_microgeographic_2014}{2014}); \protect\hyperlink{ref-lescak_evolution_2015}{Lescak et al.} (\protect\hyperlink{ref-lescak_evolution_2015}{2015}); \protect\hyperlink{ref-barrett_linking_2019}{Barrett et al.} (\protect\hyperlink{ref-barrett_linking_2019}{2019}); \protect\hyperlink{ref-divittorio_natural_2020}{DiVittorio et al.} (\protect\hyperlink{ref-divittorio_natural_2020}{2020})) likely contribute to an availability bias that leads many to perceive that local adaptation is ubiquitous. However, systematic meta-analyses reveal that local adaptation is often weak or nonexistent (\protect\hyperlink{ref-leimu_meta-analysis_2008}{Leimu and Fischer, 2008}; \protect\hyperlink{ref-hereford_quantitative_2009}{Hereford, 2009}; \protect\hyperlink{ref-brady_understanding_2019}{Brady et al., 2019}) and may be getting weaker because of anthropogenic climate change (\protect\hyperlink{ref-bontrager_climate_2020}{Bontrager et al., 2020}). Populations may not be locally adapted if differential selection is weak relative to migration or drift or they lack heritable variation in traits under selection. Alternatively, recent anthropogenic climate change may have erased the signature of local adaptation by causing some foreign populations to have higher fitness than local populations (\protect\hyperlink{ref-bontrager_climate_2020}{Bontrager et al., 2020}). For example, in the alpine plant \emph{Boechera strica}, simulated climate change (mimicking early snow melt) favors populations from lower elevations and snow addition restores the relative fitness advantage of high-elevation populations (\protect\hyperlink{ref-anderson_climate_2020}{Anderson and Wadgymar, 2020}). Understanding what causes trait variation among populations within species, whether adaptive or not, will help biologists better predict how populations will respond to environmental change.

Phenotypic clines are commonly interpreted as evidence for local adaptation, but nonadaptive processes also generate similar patterns (\protect\hyperlink{ref-endler_geographic_1977}{Endler, 1977}). A phenotypic cline is a correlation between a heritable character and the environment or a proxy for environmental gradients such as latitude or elevation (\protect\hyperlink{ref-huxley_clines_1938}{Huxley, 1938}). Clines may be caused by either abrupt or gradual environmental change. For example, heavy metal contamination of soil around mines is an abrupt change from the surrounding pasture soil. There is an abrupt change in the heavy metal tolerance of \emph{Anthoxanthum odoratum} populations across the mine boundary, but other traits under weaker or correlated selection vary gradually over space (\protect\hyperlink{ref-antonovics_evolution_1970}{Antonovics and Bradshaw, 1970}). Other clines track gradual environmental change. Latitudinal clines in the size of \emph{Drosophila subobscura} flies have evolved multiple times and most likely track gradual variation in temperature and/or phenology. But clines are not necessarily adaptive. When there is genetic isolation by distance (\protect\hyperlink{ref-wright_isolation_1943}{Wright, 1943}), it is possible for phenotypic differences to track genetic differences throughout the species range (\protect\hyperlink{ref-vasemagi_adaptive_2006}{Vasemägi, 2006}). In practice, nonadaptive clines are difficult to demonstrate because it is hard to reject adaptive explanations. The proportion of cyanogenesis varies clinally in white clover (\emph{Trifolium repens}) but field experiments have not observed spatially varying selection which could explain this cline (\protect\hyperlink{ref-wright_genetic_2021}{Wright et al., 2021}). Hence, spatial variation in cyanogenesis may be nonadaptive or the design of the experiment may have missed selection occurring during germination and early development. When species expand their range, such as during biological invasions, nonadaptive clines can readily evolve because of multiple introductions and serial founder events (\protect\hyperlink{ref-colautti_contemporary_2015}{Colautti and Lau, 2015}). Even parallel clines, which are usually considered strong evidence of natural selection, can result from nonadaptive processes when there is epistasis and consistent spatial variation in the strength of genetic drift (\protect\hyperlink{ref-santangelo_modern_2018}{Santangelo et al., 2018}). Evolutionary ecological experiments such as reciprocal transplants are needed to provide additional evidence to distinguish between adaptive and nonadaptive hypotheses for clines (\protect\hyperlink{ref-wadgymar_integrating_2017}{Wadgymar, Daws, et al., 2017}).

There are several experimental methods for measuring the agents and strength of selection to different environments under natural conditions (\protect\hyperlink{ref-wadgymar_identifying_2017}{Wadgymar, Lowry, et al., 2017}), but here we focus on differential selection. One key prediction is that the direction of selection and position of the phenotypic optimum will differ between environments. If genetic variation in a trait \emph{causes} fitness to vary, then genotypic selection analyses (\protect\hyperlink{ref-rausher_measurement_1992}{Rausher, 1992}) should find that genotypes with high trait expression are favored in one environment but disfavored in another. For example, selection favors higher seed dormancy in southern environments of \emph{Arabidopsis thaliana} and lower dormancy in northern environments, presumably to optimally time seedling establishment with the onset of favorable summer conditions (\protect\hyperlink{ref-postma_early_2016}{Postma and Ågren, 2016}). Populations can also simultaneously differ as a result of both adaptive and non-adaptive processes. When locally adapted populations differ in traits that do not themselves confer local adaptation, we expect population-level covariation between divergent traits and fitness but do not expect within-population variation to correlate with fitness. Hence, comparing the direction of among-genotype and among-population selection can help distinguish between phenotypic clines that are important for local adaptation and those which may have evolved nonadaptively.

In this study, we used a reciprocal transplant to test whether a cline in \emph{Mimulus cardinalis} germination rate could be locally adaptive in terms of early survival between seedling establishment and flowering. Because seeds of \emph{M. cardinalis} do not require cold or warm stratification to break dormancy, they will germinate whenever conditions are sufficiently warm and moist. Within this germination niche, there is genetic variation for germination time that covaries with latitude and other traits associated with fast-slow growth continuum (\protect\hyperlink{ref-muir_grow_2017}{Muir and Angert, 2017}; \protect\hyperlink{ref-sheth_demographic_2018}{Sheth and Angert, 2018}; \protect\hyperlink{ref-nelson_quantitative_2021}{Nelson et al., 2021}). In greenhouse conditions, seeds from the southern end of the range (San Diego County, CA) germinate in 6-7 days; seeds from the northern end of the range (Lane County, OR) germinate in 9-10 days. The adaptive hypothesis is that faster germination increases early survival in the southern end of the range and slower germination increases early survival at the northern end. We would predict either stabilizing genotypic selection for different optimal germination rates or opposing signs of directional genotypic selection. If the cline is maintained by nonadaptive forces then germination rate should be uncorrelated with genotypic fitness (even if it is correlated with population mean fitness) or the direction of selection should be the same in both parts of the range. A limitation of this study is that we cannot measure selection on seedling establishment since that occurred in the greenhouse (see \protect\hyperlink{materials-and-methods}{Materials and Methods}) or later survival and fecundity (these fitness components will be analyzed in a follow-up paper).

Germination and early survival are potentially important for local adaptation but often overlooked compared to later life stages. Failure to measure seedling establishment and early survival can bias estimates of selection on traits that only express at later life stages, such as flower size (\protect\hyperlink{ref-mojica_viability_2010}{Mojica and Kelly, 2010}). In the herb \emph{Arabidopsis thaliana}, quantitative differences in germination timing contributed to local adaptation and are maintained by a balance of selection on seedling establishment, survival, and fecundity, but the optimal timing varies in northern and southern environments (\protect\hyperlink{ref-postma_early_2016}{Postma and Ågren, 2016}, \protect\hyperlink{ref-postma_among-year_2018}{2018}). A challenge of measuring genotypic selection on germination rate is that quantitative genetic theory and analysis are built around Gaussian distributions (\protect\hyperlink{ref-lynch_genetics_1998}{Lynch and Walsh, 1998}; but see \protect\hyperlink{ref-de_villemereuil_general_2016}{Villemereuil et al., 2016}), but the distribution of times until germination are most likely not Gaussian. Note that time to germination is the inverse of germination rate. The time until germination can be modeled as a waiting time, which are bounded at zero and usually right skewed. Waiting times are distributed exponentially if the probability of germination is constant through time, but this unlikely to be true. Rather, the probability of germination almost certainly increases through time once conditions are conducive. Hence, more complex waiting time distributions should be seen. A second challenge is that waiting time is usually modeled as a continuous trait, but is observed at discrete intervals (e.g.~once per day). If the average time to germination is sufficiently long relative to the interval between observations, then a continuous-time model may be sufficient. However, when time to germination is several days, as with \emph{M. cardinalis}, then accurately modeling the trait may need to account for the discrete observation process. Hence, investigating spatially variable selection on germination time is important for expanding our knowledge of local adaptation and phenotypic clines for poorly studied life stages with challenging statistical properties.

@@

\emph{Mimulus cardinalis} (\(=\)\emph{Erythranthe cardinalis} (\protect\hyperlink{ref-lowry_case_2019}{Lowry et al., 2019})) is a perennial forb native to the Western US (California and Oregon). It is predominantly outcrossing, self-compatible, and hummingbird-pollinated. We used five focal populations from throughout the geographic range of \emph{M. cardinalis} (Table \ref{tab:focal_pops}). These five focal populations are a subset of those we used previously to identify a cline in germination and other traits (\protect\hyperlink{ref-muir_grow_2017}{Muir and Angert, 2017}). Seeds were collected in the field from mature, undehisced fruits left in open coin envelopes for 2-4 weeks to dry, then stored at room temperature. To control for maternal effects, we grew a large number of field-collected seeds in the greenhouse and generated seed families for this experiment by hand-pollinating individuals using the breeding design described in the next section.

\begin{longtable}[]{@{}lccc@{}}
\caption{\label{tab:focal_pops} Source populations, including the name of the drainage where the seeds were collected, the latitude, longitude, and elevation in meters above sea level (mas).}\tabularnewline
\toprule
Name & Latitude & Longitude & Elevation (mas) \\
\midrule
\endfirsthead
\toprule
Name & Latitude & Longitude & Elevation (mas) \\
\midrule
\endhead
Sweetwater River & 32.900 & -116.585 & 1180 \\
West Fork Mojave River & 34.284 & -117.378 & 1120 \\
North Fork Middle Tule River & 36.201 & -118.759 & 926 \\
Little Jamison Creek & 39.743 & -120.704 & 1603 \\
Rock Creek & 43.374 & -122.957 & 326 \\
\bottomrule
\end{longtable}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Selection on early survival does not explain germination rate clines in Mimulus cardinalis (Phrymaceae)},
  pdfauthor={Christopher D. Muir\^{}\{1,2,*\}, Courtney L. Van Den Elzen\^{}\{1,3\}, Amy L. Angert\^{}\{1,4\}},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{geometry}
\geometry{verbose,letterpaper,margin=2.45cm}

@@

\usepackage{lineno}
\linenumbers % toggle on for linenumbers

@@

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

@@

\setstretch{1.5}
\(^{1}\) Department of Botany, University of British Columbia, Vancouver, BC V6T 1Z4, Canada

@@

Manuscript received \_\_\_\_\_\_\_; revision accepted \_\_\_\_\_\_\_.

\emph{Short title}: Selection on germination in \emph{Mimulus cardinalis}

@@

\textbf{Conclusions} Southern populations have higher early survival but this is not caused by faster germination. This pattern is consistent with nonadaptive forces driving the phenotypic cline in germination rate, but future work will need to assess whether there is selection at other life stages. This statistical framework should help expand quantitative genetic analyses for other waiting-time traits.

@@

Populations within species differ in ecologically important traits that often result from adaptation to different local environments (\protect\hyperlink{ref-turesson_genotypic_1922}{Turesson, 1922}; \protect\hyperlink{ref-clausen_experimental_1948}{Clausen et al., 1948}). However, it is still rare that we understand the traits and selective agents underlying local adaptation (\protect\hyperlink{ref-wadgymar_identifying_2017}{Wadgymar, Lowry, et al., 2017}). The classic signature of local adaptation is crossing reaction norms for fitness measured in a reciprocal transplant experiment (\protect\hyperlink{ref-kawecki_conceptual_2004}{Kawecki and Ebert, 2004}; \protect\hyperlink{ref-johnson_reciprocal_2021}{Johnson et al., 2021}), such that populations have higher relative fitness in their local environment but lower relative fitness in a foreign environment. For example, annual inland ecotypes of the yellow Monkeyflower \emph{Mimulus guttatus} have higher relative fitness in seasonally dry sites compared to perennial coastal ecotypes that are adapted to year-round water availability. The relative fitness of coastal and inland ecotypes is reversed on the coast because of life history and other genetic differences between ecotypes (\protect\hyperlink{ref-lowry_widespread_2010}{Lowry and Willis, 2010}). The prevalence of local adaptation implies that selection varies spatially, populations have heritable variation in fitness, and selection is stronger than migration or drift. Exceptional cases of local adaptation over short temporal or spatial scales (e.g. \protect\hyperlink{ref-wright_indirect_2013}{Wright et al., 2013}; \protect\hyperlink{ref-grant_40_2014}{Grant and Grant, 2014}; \protect\hyperlink{ref-richardson_microgeographic_2014}{Richardson et al., 2014}; \protect\hyperlink{ref-lescak_evolution_2015}{Lescak et al., 2015}; \protect\hyperlink{ref-barrett_linking_2019}{Barrett et al., 2019}; \protect\hyperlink{ref-divittorio_natural_2020}{DiVittorio et al., 2020}) likely contribute to an availability bias that leads many to perceive that local adaptation is ubiquitous. However, systematic meta-analyses reveal that local adaptation is often weak or nonexistent (\protect\hyperlink{ref-leimu_meta-analysis_2008}{Leimu and Fischer, 2008}; \protect\hyperlink{ref-hereford_quantitative_2009}{Hereford, 2009}; \protect\hyperlink{ref-brady_understanding_2019}{Brady et al., 2019}) and may be getting weaker because of anthropogenic climate change (\protect\hyperlink{ref-bontrager_climate_2020}{Bontrager et al., 2020}). Populations may not be locally adapted if differential selection is weak relative to migration or drift or they lack heritable variation in traits under selection. Alternatively, recent anthropogenic climate change may have erased the signature of local adaptation by causing some foreign populations to have higher fitness than local populations (\protect\hyperlink{ref-bontrager_climate_2020}{Bontrager et al., 2020}). For example, in the alpine plant \emph{Boechera strica}, simulated climate change (mimicking early snow melt) favors populations from lower elevations and snow addition restores the relative fitness advantage of high-elevation populations (\protect\hyperlink{ref-anderson_climate_2020}{Anderson and Wadgymar, 2020}). Understanding what causes trait variation among populations within species, whether adaptive or not, will help biologists better predict how populations will respond to environmental change.

Phenotypic clines are commonly interpreted as evidence for local adaptation, but nonadaptive processes also generate similar patterns (\protect\hyperlink{ref-endler_geographic_1977}{Endler, 1977}). A phenotypic cline is a correlation between a heritable character and the environment or a proxy for environmental gradients such as latitude or elevation (\protect\hyperlink{ref-huxley_clines_1938}{Huxley, 1938}). Clines may be caused by either abrupt or gradual environmental change. For example, heavy metal contamination of soil around mines is an abrupt change from the surrounding pasture soil. A similarly abrupt change in the heavy metal tolerance of \emph{Anthoxanthum odoratum} occurs across the mine boundary, but other traits under weaker or correlated selection vary gradually over space (\protect\hyperlink{ref-antonovics_evolution_1970}{Antonovics and Bradshaw, 1970}). Other clines track gradual environmental change. Latitudinal clines in the size of \emph{Drosophila subobscura} flies have evolved multiple times and most likely track gradual variation in temperature and/or phenology (\protect\hyperlink{ref-huey_rapid_2000}{Huey et al., 2000}). But clines are not necessarily adaptive. When there is genetic isolation by distance (\protect\hyperlink{ref-wright_isolation_1943}{Wright, 1943}), it is possible for phenotypic differences to correlate with genetic differences throughout the species range (\protect\hyperlink{ref-vasemagi_adaptive_2006}{Vasemägi, 2006}). In practice, nonadaptive clines are difficult to demonstrate because it is hard to reject adaptive explanations. The proportion of cyanogenesis varies clinally in white clover (\emph{Trifolium repens}) but field experiments have not observed spatially varying selection that could explain this cline (\protect\hyperlink{ref-wright_genetic_2021}{Wright et al., 2021}). Hence, spatial variation in cyanogenesis may be nonadaptive or the design of the experiment may have missed selection occurring during germination and early development. When species expand their range, such as during biological invasions, nonadaptive clines can readily evolve because of multiple introductions and serial founder events (\protect\hyperlink{ref-colautti_contemporary_2015}{Colautti and Lau, 2015}). Even parallel clines, which are usually considered strong evidence of natural selection, can result from nonadaptive processes when there is epistasis and consistent spatial variation in the strength of genetic drift (\protect\hyperlink{ref-santangelo_modern_2018}{Santangelo et al., 2018}). Evolutionary ecological experiments such as reciprocal transplants are needed to provide additional evidence to distinguish between adaptive and nonadaptive hypotheses for clines (\protect\hyperlink{ref-wadgymar_integrating_2017}{Wadgymar, Daws, et al., 2017}).

There are several experimental methods for measuring the agents and strength of selection to different environments under natural conditions (\protect\hyperlink{ref-wadgymar_identifying_2017}{Wadgymar, Lowry, et al., 2017}), but here we focus on differential selection. One key prediction is that the direction of selection and position of the phenotypic optimum will differ between environments. If genetic variation in a trait \emph{causes} fitness to vary, then genotypic selection analyses (\protect\hyperlink{ref-rausher_measurement_1992}{Rausher, 1992}) should find that genotypes with high trait expression are favored in one environment but disfavored in another. For example, selection favors higher seed dormancy in southern environments of \emph{Arabidopsis thaliana} and lower dormancy in northern environments, presumably to optimally time seedling establishment with the onset of favorable summer conditions (\protect\hyperlink{ref-postma_early_2016}{Postma and Ågren, 2016}). Populations can also simultaneously differ as a result of both adaptive and non-adaptive processes. When locally adapted populations differ in traits that do not themselves confer local adaptation, we expect population-level covariation between divergent traits and fitness but do not expect within-population variation to correlate with fitness. Hence, comparing the direction of among-genotype and among-population selection can help distinguish between phenotypic clines that are important for local adaptation and those that may have evolved nonadaptively.

In this study, we used a reciprocal transplant to test whether a cline in \emph{Mimulus cardinalis} germination rate could be locally adaptive in terms of early survival between seedling establishment and flowering. Because seeds of \emph{M. cardinalis} do not require cold or warm stratification to break dormancy, they will germinate whenever conditions are sufficiently warm and moist. Within this germination niche, there is genetic variation for germination time that covaries with latitude and other traits associated with a fast-slow growth continuum (\protect\hyperlink{ref-muir_grow_2017}{Muir and Angert, 2017}; \protect\hyperlink{ref-sheth_demographic_2018}{Sheth and Angert, 2018}; \protect\hyperlink{ref-nelson_quantitative_2021}{Nelson et al., 2021}). In greenhouse conditions, seeds from the southern end of the range (San Diego County, CA) germinate in \(6 \textendash 7\) days; seeds from the northern end of the range (Lane County, OR) germinate in \(9 \textendash 10\) days. The growing season is typically longer in the southern end of the range because of reduced or nonexistent snowpack, but for the same reason southern populations may be more susceptible to late-season drought (\protect\hyperlink{ref-sheth_demographic_2018}{Sheth and Angert, 2018}). The adaptive hypothesis is that faster germination increases early survival in the southern end of the range and slower germination increases early survival at the northern end. We would predict either stabilizing genotypic selection for different optimal germination rates or opposing signs of directional genotypic selection. If the cline is maintained by nonadaptive forces then germination rate should be uncorrelated with genotypic fitness (even if it is correlated with population mean fitness) or the direction of selection should be the same in both parts of the range. A limitation of this study is that we cannot measure selection on seedling establishment since that occurred in the greenhouse (see \protect\hyperlink{materials-and-methods}{Materials and Methods}) or later survival and fecundity (these fitness components will be analyzed in a follow-up paper).

Germination and early survival are potentially important for local adaptation but often overlooked compared to later life stages. Failure to measure seedling establishment and early survival can bias estimates of selection on traits that only express at later life stages, such as flower size (\protect\hyperlink{ref-mojica_viability_2010}{Mojica and Kelly, 2010}). In the herb \emph{Arabidopsis thaliana}, quantitative differences in germination timing contributed to local adaptation and are maintained by a balance of selection on seedling establishment, survival, and fecundity, but the optimal timing varies in northern and southern environments (\protect\hyperlink{ref-postma_early_2016}{Postma and Ågren, 2016}, \protect\hyperlink{ref-postma_among-year_2018}{2018}). A challenge of measuring genotypic selection on germination rate is that quantitative genetic theory and analysis are built around Gaussian distributions (\protect\hyperlink{ref-lynch_genetics_1998}{Lynch and Walsh, 1998}; but see \protect\hyperlink{ref-de_villemereuil_general_2016}{Villemereuil et al., 2016}), but the distribution of times until germination are most likely not Gaussian. Although it is convenient to express germination as a rate (events per time), this and other ``waiting time'' traits are measured as the time until an event, the inverse of germination rate. Throughout this paper we distinguish between germination rate as the trait of interest which we estimate by measuring the time to germination. The time until germination (or another event) can be modeled as a waiting time, which is bounded at zero and usually right skewed. Waiting times are distributed exponentially if the probability of germination is constant through time, but this unlikely to be true. Rather, the probability of germination almost certainly increases through time once conditions are conducive. Hence, more complex waiting time distributions should occur in nature. A second challenge is that waiting time is usually modeled as a continuous trait, but is observed at discrete intervals (e.g.~once per day). If the average time to germination is sufficiently long relative to the interval between observations, then a continuous-time approximation may be sufficient. However, when time to germination is several days, as with \emph{M. cardinalis}, then accurately modeling the trait may need to account for the discrete observation process. Here we show that non-Gaussian distributions and measurement process can be incorporated into and improve analysis of time to germination. This approach should be useful, with some customization, to many other traits and biological systems. Hence, investigating spatially variable selection on germination time is important for expanding our knowledge of local adaptation and phenotypic clines for poorly studied life stages with challenging statistical properties.

@@

\emph{Mimulus cardinalis} (\(=\)\emph{Erythranthe cardinalis} \protect\hyperlink{ref-lowry_case_2019}{Lowry et al., 2019}) is a perennial forb native to the Western US (California and Oregon). It is self-compatible, predominantly outcrossing, and hummingbird pollinated. We used five focal populations from throughout the geographic range of \emph{M. cardinalis} (Table \ref{tab:focal_pops}; Fig. \ref{fig:range-map}). These five focal populations are a subset of those we used previously to identify a cline in germination and other traits (\protect\hyperlink{ref-muir_grow_2017}{Muir and Angert, 2017}). Seeds were collected in the field from mature, undehisced fruits left in open coin envelopes for \(2-4\) weeks to dry, then stored at room temperature. To minimize maternal effects, we grew a large number of field-collected seeds in the greenhouse and generated seed families for this experiment by hand-pollinating individuals using the breeding design described in the next section.
